A thermodynamic view of dusty protoplanetary disks

### 2017-07-01 paper.tex
deleted original section two; remixed original section 5 and 6 (now 4 and 5) so the stability discussion goes from general to special, from not needing explicit background to explicit dust distributions

\section{Introduction}
%si, sgi as examples
Protoplanetary disks are comprised of a mixture of gas and dust
\citep{chiang10}. Although the overall dust content is small
($\sim 1\%$ by mass), their presence can have profound impacts on the 
gas dynamics and vice versa. Dust-gas friction introduces phenomena
that are absent in pure gas disks. Important examples include the
streaming instability 
\citep[SI,][]{youdin05a,youdin07b,johansen07}, secular gravitational
instabilities \citep[SGI,][]{ward00,youdin11,michikoshi12,takahashi14} and
Kelvin-Helmholtz instabilities
\citep{goldreich73, chiang08,barranco09,lee10}. These instabilities 
can trigger turbulence, and (in different regimes) promote or inhibit 
planetesimal formation. 
More recently, new 
dusty instabilities have appeared in numerical
simulations \citep{loren15,loren16,lamb16}, but their physical underpinnings are not yet well
understood.

%particles 
Current state-of-the-art models of dusty protoplanetary
disks directly simulate gas dynamics coupled to explicit
Lagragian dust particles
\citep{johansen06, nelson10,bai10,yang14,zhu14,gibbons15,simon16,baruteau16}. 
The equation of motion for each particle is solved 
including dust-gas drag; the strength of which is measured 
by a stopping time $\tstop$ --- the decay timescale for the 
relative velocity between gas and dust. Thus small $\tstop$
corresponds to tightly-coupled particles.

@@

The advantage of this reformulation arises in the tight coupling limit of  $\tstop \ll 1/\Omega$, where the orbital frequency $\Omega$ sets the characteristic timescale in many disk dynamics problems.
Models with two fluids or with Lagrangian solids are numerically difficult to evaluate in this regime, due to stiff drag forces.  
In the reformulated equations, however, the center of mass motion does not experience drag forces and thus is not stiff.  Moreover the relative motion satisfies the terminal-velocity approximation and can be eliminated from the equations.  This single fluid framework for a well-coupled dust-gas mixture facilitates numerical calculations \citep{price15}. and also simplifies analytic calculations \citep{youdin05a,jacquet11}, which is useful for gaining insight.

@@

The purpose of our physical and mathematical parallel between
dusty gas and standard hydrodynamics is to gain a deeper understanding of the dynamics
and stability of dusty disks.  
We can apply established methods to generalize previous stability 
results for pure gas dynamics to dusty disks.  
We show that strictly isothermal protoplanetary disks with
perfectly-coupled dust are generally 
stable to axisymmetric perturbations.%, regardless of how  thin the dust
%layer is. 
{\bf should we mention vertically well mixed case here?} 
We put forward a thermodynamical interpretation of 
dust-gas drag instabilities, and apply it to the streaming instability.   
We also study the effect of dust-loading on the vertical shear
instability previously considered in pure gas disks
\citep[VSI,][]{nelson13,lin15,barker15}.

@@

\section{Dust-free modeling of dusty protoplanetary disks}\label{key_idea}
{\bf is this section needed?} 
The key concept we introduce in this work is the description 
of dusty protoplanetary disks \emph{without} explicit dust 
equations. This is best illustrated by writing down our model   
equations for a mixture of isothermal gas and small 
dust particles: 
\begin{align}  
  &\frac{\p \rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right)= 0, \label{mass0}\\ 
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} = -
  \frac{1}{\rho}\nabla  P - \nabla \Phi_*,\label{mom0}\\ 
 & \frac{\p P}{\p t} + \bm{v}\cdot\nabla P 
  = - P \nabla\cdot\bm{v} - \nabla\cdot\bm{\mathcal{F}}
%                + \mathcal{C}(\rho, P; \tstop)
\label{en0}, 
\end{align}
where $\rho=\rhod + \rhog$ is the total density, $P$ is the pressure,
$\Phi_*$ is the stellar gravitational potential,   
and $\bm{\mathcal{F}}=~\bm{\mathcal{F}}(\rho, P; \tstop)$ model effects of dust-gas drag, explained below. 
%$\mathcal{C}$ represent effects of dust-gas drag, which
%is paremterized by the stopping time $\tstop$ ($\mathcal{C}=0$ for
%$\tstop=0$). 
Similar equations for dusty-gas are given by
\cite{price15}.

Eq. \ref{mass0}---\ref{en0} appear identical to pure gas
dynamics, but they physically represents a two-phase, 
dusty gas. The mixture obeys mass and momentum conservation as usual,
but also an energy-like equation, even though the system is
isothermal by assumption. The term $-\bm{\mathcal{F}}\propto \tstop
\nabla P$ is not a true heat flux. Instead, it reflects 
particle flux towards pressure maximum due to dust-gas drag 
\citep{weidenschilling77}. The back-reaction onto the gas then changes
the pressure distribution.

%weidenschilling7

Since the above equations are similar to ideal gas dynamics, it 
means that many of the results derived for ideal hydrodynamics can be  
generalized to dusty disks without additional effort. This includes 
the important consequence of phase lags between density and pressure,
discussed next.

>>>>

\section{Introduction}
%si, sgi as examples
Protoplanetary disks are comprised of a mixture of gas and dust
\citep{chiang10}. Although the overall dust content is small
($\sim 1\%$ by mass), their presence can have profound impacts on the 
gas dynamics and vice versa. Dust-gas friction introduces phenomena
that are absent in pure gas disks. Important examples include the
streaming instability 
\citep[SI,][]{youdin05a,youdin07b,johansen07}, secular gravitational
instabilities \citep[SGI,][]{ward00,youdin11,michikoshi12,takahashi14} and
Kelvin-Helmholtz instabilities
\citep{goldreich1973, chiang08,barranco09,lee10}. These instabilities 
can trigger turbulence, and (in different regimes) promote or inhibit 
planetesimal formation. 
More recently, new 
dusty instabilities have appeared in numerical
simulations \citep{loren15,loren16,lamb16}, but their physical underpinnings are not yet well
understood.

%particles 
Current state-of-the-art models of dusty protoplanetary
disks directly simulate gas dynamics coupled to explicit
Lagragian dust particles
\citep{johansen2006, nelson10,bai10,yang14,zhu14,gibbons15,simon16,baruteau16}. 
The equation of motion for each particle is solved 
including dust-gas drag; the strength of which is measured 
by a stopping time $\tstop$ --- the decay timescale for the 
relative velocity between gas and dust. Thus small $\tstop$
corresponds to tightly-coupled particles.

@@

The advantage of this reformulation arises in the tight coupling limit of  $\tstop \ll 1/\Omega$, where the orbital frequency $\Omega$ sets the characteristic timescale in many disk dynamics problems.
Models with two fluids or with Lagrangian solids are numerically difficult to evaluate in this regime, due to stiff drag forces.  
In the reformulated equations, however, the center of mass motion does
not experience drag forces and thus is not stiff.  Moreover,
for $\tstop\Omega\ll 1$ the relative motion satisfies the terminal-velocity approximation and can be eliminated from the equations.  This single fluid framework for a well-coupled dust-gas mixture facilitates numerical calculations \citep{price15}; and also simplifies analytic calculations \citep{youdin05a,jacquet11}, which is useful for gaining insight.

@@

The purpose of our physical and mathematical parallel between
dusty gas and standard hydrodynamics is to gain a deeper understanding of the dynamics
and stability of dusty disks.  
We can apply established methods to generalize previous stability 
results for pure gas dynamics to dusty disks.  
We show that strictly isothermal protoplanetary disks with
perfectly-coupled dust are generally 
stable to axisymmetric perturbations, unless the 
dust-to-gas ratio varies more rapidly in radius than in height.  
%, regardless of how  thin the dust
%layer is. 
%{\bf should we mention vertically well mixed case here?} 
We put forward a thermodynamical interpretation of 
dust-gas drag instabilities, and apply it to the streaming instability.   
We also study the effect of dust-loading on the vertical shear
instability previously considered in pure gas disks
\citep[VSI,][]{nelson13,lin15,barker15}.

@@

% \section{Dust-free modeling of dusty protoplanetary disks}\label{key_idea}
% {\bf is this section needed?} 
% The key concept we introduce in this work is the description 
% of dusty protoplanetary disks \emph{without} explicit dust 
% equations. This is best illustrated by writing down our model   
% equations for a mixture of isothermal gas and small 
% dust particles: 
% \begin{align}  
%   &\frac{\p \rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right)= 0, \label{mass0}\\ 
%   &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} = -
%   \frac{1}{\rho}\nabla  P - \nabla \Phi_*,\label{mom0}\\ 
%  & \frac{\p P}{\p t} + \bm{v}\cdot\nabla P 
%   = - P \nabla\cdot\bm{v} - \nabla\cdot\bm{\mathcal{F}}
% %                + \mathcal{C}(\rho, P; \tstop)
% \label{en0}, 
% \end{align}
% where $\rho=\rhod + \rhog$ is the total density, $P$ is the pressure,
% $\Phi_*$ is the stellar gravitational potential,   
% and $\bm{\mathcal{F}}=~\bm{\mathcal{F}}(\rho, P; \tstop)$ model effects of dust-gas drag, explained below. 
% %$\mathcal{C}$ represent effects of dust-gas drag, which
% %is paremterized by the stopping time $\tstop$ ($\mathcal{C}=0$ for
% %$\tstop=0$). 
% Similar equations for dusty-gas are given by
% \cite{price15}.

% Eq. \ref{mass0}---\ref{en0} appear identical to pure gas
% dynamics, but they physically represents a two-phase, 
% dusty gas. The mixture obeys mass and momentum conservation as usual,
% but also an energy-like equation, even though the system is
% isothermal by assumption. The term $-\bm{\mathcal{F}}\propto \tstop
% \nabla P$ is not a true heat flux. Instead, it reflects 
% particle flux towards pressure maximum due to dust-gas drag 
% \citep{weidenschilling77}. The back-reaction onto the gas then changes
% the pressure distribution.

% %weidenschilling7

% Since the above equations are similar to ideal gas dynamics, it 
% means that many of the results derived for ideal hydrodynamics can be  
% generalized to dusty disks without additional effort. This includes 
% the important consequence of phase lags between density and pressure,
% discussed next.
